
# Filtering, Subsetting, and Summarising using dplyr

## Prerequisites

The prerequisites for this tutorial are `tidyverse` and `rclimateca`. If these packages aren't installed, you'll have to install them using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("rclimateca")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
library(rclimateca)
```

Finally, you will need to obtain the sample data using the `getClimateData()` function in the **rclimateca** package. Copy/paste the statement below to load the sample data for this tutorial.

```{r}
climate_data <- getClimateData(c(27141, 6354), nicenames = TRUE)
```

## Viewing a Data Frame

In RStudio's "Environment" tab (usually at the top right of the screen), you should see a variable called `climate_data` in the list. You can inspect it by clicking on the variable name, after which a tab will appear displaying the contents of the variable you just loaded. Clicking the little arrow to the left of the name will display the structure of the data frame, including the column names and some sample values. You can also do both of these things using the R commands `View()` and `str()`, respectively. Also useful is the `head()` function, which will display the first few rows of a data frame.

```{r, eval=FALSE}
View(climate_data) # will display a graphic table browser
```

```{r}
str(climate_data) # will display a text summary of the object
```

```{r}
head(climate_data)
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial. Each row contains a number of parameters that are available on a monthly basis from two Environment Canada climate stations (Kentville Agricultural Research Station is 27141; Greenwood Station is 6354). The `stationid` column identifies where the values were measured, the `year` and `month` column identify when the values were measured, and the rest of the columns contain the measured values. For each measured value column, there is a `flag` column that gives additional information about the measurement (for the most part, we will ignore these columns). The only column names that are slightly cryptic are `extrmaxtemp` and `extrmintemp`, which are the extreme maximum and minimum temperatures measured in that month, respectively.

## Selecting Columns

One way to subset `climate_data` is to subset by column, for which we will use the `select()` function. For example, we may only be interested in the mean temperature information, represented by the columns `stationid`, `year`, `month`, and `meantemp`.

```{r}
mean_temp_data <- select(climate_data, stationid, year, month, meantemp)
```

The first argument to the `select()` function is the original data frame (in this case, `climate_data`), and the remaining arguments are the names of the columns to be selected. To select the `stationid`, `year`, `month`, `meantemp` and `totalprecip` columns, you would use the following R command:

```{r}
temp_precip_data <- select(climate_data, stationid, year, month, meantemp, totalprecip)
```

### Excercises

- Use `View()`, `str()`, and `head()` to preview the two data frames we just created. Do they have the columns you would expect?
- Use `select()` to select `stationid`, `year`, `month`, and all of the columns containing temperature values, and assign it to the variable `temp_data`.

```{r, include = FALSE, eval = FALSE}
# using View, str, and head
View(mean_temp_data); str(mean_temp_data); head(mean_temp_data)
View(temp_precip_data); str(temp_precip_data); head(temp_precip_data)

# selecting all the temperature variables
temp_data <- select(climate_data, meanmaxtemp, meanmintemp, meantemp,
                    extrmaxtemp, extrmintemp)
```

## Filtering Rows

Another way to subset `climate_data` is by filtering rows using column values, similar to the filter feature in Microsoft Excel. This is done using the `filter()` function. For example, we may only be interested in July temperature for the two stations.

```{r}
july_data <- filter(climate_data, month == 7)
```

Just like `select()`, the first argument to `filter()` is the original data frame, and the subsequent arguments are the conditions that each row must satisfy in order to be included in the output. Column values are referred to by the column name (in the above example, `month`), so to include all rows where the value in the `month` column is 7, we use `month == 7`. Passing multiple conditions means each row must satisfy all of the conditions, such that to obtain the data for July of 1999, we can use the following call to `filter()`:

```{r}
july_1999_data <- filter(climate_data, month == 7, year == 1999)
```

It is very important that there are two equals signs within `filter()`! The `==` operator tests for equality (e.g. `(2 + 2) == 4`), whereas the `=` operator assigns a value or passes a named argument to a function, which is not what you're trying to do within `filter()`. Other common operators that are useful within filter are `!=` (not equal to), `>` (greater than), `<` (less than), `>=` (greater than or equal to), `<=` (less than or equal to), and `%in%` (tests if the value is one of several values). Using these, we could find out which observations had mean temperatures that were below freezing:

```{r}
freezing_observations <- filter(climate_data, meantemp < 0)
```

We could also find which observations occurred during the summer months (May, June, July, or August):

```{r}
summer_data <- filter(climate_data, month %in% c(5, 6, 7, 8))
```

### Exercises

- Use `View()`, `str()`, and `head()` to preview the data frames we just created. Do they have the rows you would expect?
- Use `filter()` to find observations from the month of December where the mean temperature was above freezing.
- Are there any observations from the month of January where the mean temperature was below freezing?
- Filter `climate_data` to include only observations from the months of December, January, February, and March and assign it to a variable name of your choosing.

```{r, include = FALSE, eval = FALSE}
# View, str, and head as above
filter(climate_data, month == 12, meantemp > 0)
filter(climate_data, month == 1, meantemp > 0)
winter_data <- filter(climate_data, month %in% c(12, 1, 2, 3))
```

## Selecting and Filtering

Often we need to use both `select()` and `filter()` to obtain the desired subset of a data frame. To do this, we need to pass the result of `select()` to `filter()`, or the result of `filter()` to select. For example, we could create a data frame of mean temperature observations from the month of July one of two ways (you'll recall that we selected temperature columns in the data frame `mean_temp_data`, and we filtered for the month of July in the data frame `july_data`):

```{r}
july_temp <- filter(mean_temp_data, month == 7)
july_temp2 <- select(july_data, stationid, year, month, meantemp)
```

### Exercises

- Inspect `july_temp` and `july_temp2` using `View()`, `str()`, and `head()`. Are they identical?
- Create a data frame of July total precipitation data and give it a variable name of your choosing. Do this by using `select()` followed by `filter()`, then using `filter()` followed by `select()`. Inspect the output to ensure the data frames are identical.

```{r, include = FALSE, eval = FALSE}
# the data frames july_temp and july_temp2 are identical
july_precip <- select(july_data, stationid, year, month, totalprecip)
precip_data <- select(climate_data, stationid, year, month, totalprecip)
july_precip2 <- filter(precip_data, month == 7)
```

## The Pipe (%>%)

There is an easier way! Instead of creating intermediary variables every time we want to subset a data frame using `select()` and `filter()`, we can use the pipe operator (`%>%`) to pass the result of one function call to another. Thus, creating our `july_temp` data frame from above becomes one line with one variable assignment instead of two.

```{r}
july_temp3 <- climate_data %>% 
  filter(month == 7) %>% 
  select(stationid, year, month, meantemp)
```

What `%>%` does is pass the left side into the first argument of the function call on the right side. Thus, `filter(climate_data, month == 7)` becomes `climate_data %>% filter(month ==7)`. When using the tidyverse family of packages, you should use the pipe as often as possible! It usually makes for more readable, less error-prone code, and reduces the number of temporary variables you create that clutter up your workspace. When using `filter()` and `select()` with other tidyverse manipulations like `arrange()`, `group_by()`, `summarise()`, and `mutate()`, the pipe becomes indispensable.

### Exercises

- Inspect `july_temp3` to ensure it is identical to `july_temp`.
- Create a data frame of July total precipitation data using `climate_data`, `filter()`, `select()`, and `%>%`. Is it identical to the data frame you created in the exercise above?

```{r, include = FALSE, eval = FALSE}
# the data frames july_temp and july_temp3 are identical
july_precip3 <- climate_data %>% 
  filter(month == 7) %>% 
  select(stationid, year, month, totalprecip)
```

## Summary

In this tutorial we introduced the use of `select()`, `filter()` and the pipe (`%>%`). These functions are the building blocks of powerful tools in the tidyverse such as `group_by()`, `summarise()`, `arrange()`, `mutate()` and many others.

