
# Creating Visualizations using ggplot

Intro

## Prerequisites

The prerequisites for this tutorial are `tidyverse` and `rclimateca`. If these packages aren't installed, you'll have to install them using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("rclimateca")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
library(rclimateca)
```

Finally, you will need to obtain the sample data using the `getClimateData()` function in the **rclimateca** package. Copy/paste the statement below to load the sample data for this tutorial.

```{r, results='hide'}
climate_data <- getClimateData(c(27141, 6354), year = 2000:2003,
                               nicenames = TRUE) %>%
  left_join(ecclimatelocs %>% select(station_name = Name, stationid = `Station ID`),
            by = "stationid")
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial. Each row contains a number of parameters that are available on a monthly basis from two Environment Canada climate stations (Kentville Agricultural Research Station is 27141; Greenwood Station is 6354). The `stationid` column identifies where the values were measured, the `year` and `month` column identify when the values were measured, and the rest of the columns contain the measured values. For each measured value column, there is a `flag` column that gives additional information about the measurement (for the most part, we will ignore these columns). The only column names that are slightly cryptic are `extrmaxtemp` and `extrmintemp`, which are the extreme maximum and minimum temperatures measured in that month, respectively.

## Using ggplot

The Grammar of Graphics (the "gg" in "ggplot") is a way of describing a graphic that is derived from data, which in R is done using the `ggplot()` function and its many friends. Unlike other plotting functions, `ggplot()` builds graphics from the data up (rather than starting with a template of a graphic and working backward). We will start with an example:

```{r}
ggplot(climate_data, aes(x = meantemp, y = totalprecip, colour = station_name)) +
  geom_point()
```


## Summary

Tutorial summary
