
# Creating Visualizations using ggplot

Intro

## Prerequisites

The prerequisites for this tutorial are `tidyverse` and `rclimateca`. If these packages aren't installed, you'll have to install them using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("rclimateca")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
library(rclimateca)
```

Finally, you will need to obtain the sample data using the `getClimateData()` function in the **rclimateca** package. Copy/paste the statement below to load the sample data for this tutorial.

Data: [canadia climate](data-raw/canada_climate.csv)

```{r, results='hide'}
# get data
climate_data <- getClimateData(c(5051, 5345, 735L, 3912, 2925, 6206, 2205, 6454, 
                                 1586, 1650, 6693, 1786, 6527),
                               nicenames = TRUE) %>%
  left_join(ecclimatelocs %>% select(station_name = Name, stationid = `Station ID`),
            by = "stationid") %>%
  select(station_name, year, month, date = parseddate, ends_with("temp"), ends_with("precip")) %>%
  set_names(., stringr::str_replace(names(.), "(temp|precip)$", "_\\1") %>%
              stringr::str_replace("(max|min)", "_\\1")) %>%
  mutate(station_name = stringr::str_to_title(station_name)) %>%
  filter(year >= 1975, year <= 2000) %>%
  as_tibble()

climate_data_moncton <- climate_data %>%
  filter(station_name == "Moncton")

# get locations
climate_locations <- ecclimatelocs %>%
  filter(`Station ID` %in% unique(c(5051, 5345, 735L, 3912, 2925, 6206, 2205, 6454, 
                                 1586, 1650, 6693, 1786, 6527))) %>%
  select(station_id = `Station ID`, station_name = Name, province = Province,
         elevation_m = `Elevation (m)`, latitude = `Latitude (Decimal Degrees)`,
         longitude = `Longitude (Decimal Degrees)`) %>%
  mutate(station_name = stringr::str_to_title(station_name)) %>%
  as_tibble()
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial. Each row contains a number of parameters that are available on a monthly basis from 13 Environment Canada climate stations, which were selected based on the longest running record available from each province/territory.

## Using ggplot

The Grammar of Graphics (the "gg" in "ggplot") is a way of describing a graphic that is derived from data, which in R is done using the `ggplot()` function and its many friends. Unlike other plotting functions, `ggplot()` builds graphics from the data up (rather than starting with a template of a graphic and working backward). We will start with an example:

```{r}
ggplot(climate_data_moncton, aes(x = date, y = mean_temp, col = factor(month))) +
  geom_line()
```

## Categorical/Grouping Variables

X, Y, Colour, Shape, Linetype, Facets

## Continuous Variables

X, Y, Colour, Size

## Summary

Tutorial summary
